---
title: "prep"
output: html_document
date: "2023-03-13"
editor_options: 
  markdown: 
    wrap: sentence
---

This is a demostration of how a long list of behavioral data (Reaction Time) is converted to a dissimilarity matrix.
"behavioral1.csv" is a dataset that consists of finger responses and its reaction time, recorded throughout the trials, by row.
There are five blocks with 210 trials in each.
The trial condition is "note" and the response time is "rt" Variable "note" is the finger movement that subjects are asked to perform.
i.e. 1000 is pressing with the little finger; 1100 for the little & ring finger; 1111 for all fingers except the thumb.
I am interested in all pairwise reaction time when going from one movement to another.
The rts are to be mapped on a dissimilarity matrix for further analysis.

### Load in the data

```{R}
library(pracma)
library(plot.matrix)

participant_n = 1                               #specifying total subject number. For demostration, we are preprocessing a single subject.
setwd("C:\\Users\\tommy\\桌面\\git")
temp= list.files()             
day_1 = temp[startsWith(temp, "behavioral1")]   #This is a multi-session study. For demostration, I selected the first session.
data = lapply(day_1, read.csv)
data[[1]][1:5,]
```

### Preprocessing

Preprocessing is done block by block.
There will be a unique matrix for each block.
This provides versatility for analysis.
One can simply average them all to check the overall effect of "note" on "rt".
Or, one can analyse the effect of "note"on "rt" within each block and compare between blocks.
i.e. fatigue effect, practice effect.
For this demostration, I will average all 5 of them.
You can see the matrix output for the first block.

```{R, warning=FALSE}
#1st block  


block = list()
for(k in 1:participant_n){
  block[[k]] = data[[k]][which(data[[k]]$trial_index==40):which(data[[k]]$trial_index==249),]
}

#get all correct responses. I am only interested in the rts of correct responses.
block_correct = list()
for(k in 1:participant_n){
  block_correct[[k]] = block[[k]][block[[k]]$correct == "TRUE",]
}

#fill in matrix
mat = matrix(NA, nrow =14, ncol=14)
colnames(mat)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(mat) = colnames(mat)
matlist = list()
for(k in 1:participant_n){
  for(i in 1:14){
    for(j in 1:14){
      mat[i,j] = mean(unlist(lapply(lapply(strsplit(gsub("\\[|\\]","",block_correct[[k]]$rt[which(block_correct[[k]]$note == colnames(mat)[j])[block_correct[[k]]$note[which(block_correct[[k]]$note == colnames(mat)[j])-1]==colnames(mat)[i]]]
      ),","), as.numeric), mean)), na.rm=T)   #This LONG line finds the rows for a specific note with correct input, for each of these rows, identify the previous row with the correct input and the specific note. 
                                              #After finding the rts for each of the conditions, then find their average (If there are multiple rts found for a specific condition).
    }
  }
}
options(width = 300)
mat
```

Do the same thing for the remaining 4 blocks.
```{R, results='hide', warning=FALSE}
#2nd block
block = list()
for(k in 1:participant_n){
  block[[k]] = data[[k]][which(data[[k]]$trial_index==251):which(data[[k]]$trial_index==461),]
}

#get all correct responses
block_correct = list()
for(k in 1:participant_n){
  block_correct[[k]] = block[[k]][block[[k]]$correct == "TRUE",]
}

#fill in matrix
mat2 = matrix(NA, nrow =14, ncol=14)
colnames(mat2)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(mat2) = colnames(mat2)
mat2list = list()
for(k in 1:participant_n){
  for(i in 1:14){
    for(j in 1:14){
      mat2[i,j] = mean(unlist(lapply(lapply(strsplit(gsub("\\[|\\]","",block_correct[[k]]$rt[which(block_correct[[k]]$note == colnames(mat2)[j])[block_correct[[k]]$note[which(block_correct[[k]]$note == colnames(mat2)[j])-1]==colnames(mat2)[i]]]
      ),","), as.numeric), mean)), na.rm=T)   
    }
  }
}

#3rd block
block = list()
for(k in 1:participant_n){
  block[[k]] = data[[k]][which(data[[k]]$trial_index==463):which(data[[k]]$trial_index==673),]
}

#get all correct responses
block_correct = list()
for(k in 1:participant_n){
  block_correct[[k]] = block[[k]][block[[k]]$correct == "TRUE",]
}

#fill in matrix
mat3 = matrix(NA, nrow =14, ncol=14)
colnames(mat3)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(mat3) = colnames(mat3)
mat3list = list()
for(k in 1:participant_n){
  for(i in 1:14){
    for(j in 1:14){
      mat3[i,j] = mean(unlist(lapply(lapply(strsplit(gsub("\\[|\\]","",block_correct[[k]]$rt[which(block_correct[[k]]$note == colnames(mat3)[j])[block_correct[[k]]$note[which(block_correct[[k]]$note == colnames(mat3)[j])-1]==colnames(mat3)[i]]]
      ),","), as.numeric), mean)), na.rm=T)   
    }
  }
}


#4th block
block = list()
for(k in 1:participant_n){
  block[[k]] = data[[k]][which(data[[k]]$trial_index==675):which(data[[k]]$trial_index==885),]
}

#get all correct responses
block_correct = list()
for(k in 1:participant_n){
  block_correct[[k]] = block[[k]][block[[k]]$correct == "TRUE",]
}

#fill in matrix
mat4 = matrix(NA, nrow =14, ncol=14)
colnames(mat4)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(mat4) = colnames(mat4)
mat4list = list()
for(k in 1:participant_n){
  for(i in 1:14){
    for(j in 1:14){
      mat4[i,j] = mean(unlist(lapply(lapply(strsplit(gsub("\\[|\\]","",block_correct[[k]]$rt[which(block_correct[[k]]$note == colnames(mat4)[j])[block_correct[[k]]$note[which(block_correct[[k]]$note == colnames(mat4)[j])-1]==colnames(mat4)[i]]]
      ),","), as.numeric), mean)), na.rm=T)   
    }
  }
}


#5th block
block = list()
for(k in 1:participant_n){
  block[[k]] = data[[k]][which(data[[k]]$trial_index==887):which(data[[k]]$trial_index==1097),]
}

#get all correct responses
block_correct = list()
for(k in 1:participant_n){
  block_correct[[k]] = block[[k]][block[[k]]$correct == "TRUE",]
}

#fill in matrix
mat5 = matrix(NA, nrow =14, ncol=14)
colnames(mat5)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(mat5) = colnames(mat5)
mat5list = list()
for(k in 1:participant_n){
  for(i in 1:14){
    for(j in 1:14){
      mat5[i,j] = mean(unlist(lapply(lapply(strsplit(gsub("\\[|\\]","",block_correct[[k]]$rt[which(block_correct[[k]]$note == colnames(mat5)[j])[block_correct[[k]]$note[which(block_correct[[k]]$note == colnames(mat5)[j])-1]==colnames(mat5)[i]]]
      ),","), as.numeric), mean)), na.rm=T)   
    }
  }
}

```

### Averaging all 5 matrices
Now that we have a matrix for every block, I can look for trend of rts over blocks. For now, I am interested in overall performance. Find average of all mats. 
```{r, fig.width= 8, fig.height= 8}
allblocks_matlist<-list(mat,mat2,mat3,mat4,mat5)
Y <- do.call(cbind, allblocks_matlist)
Y <- array(Y, dim=c(dim(allblocks_matlist[[1]]), length(allblocks_matlist)))
mean_rt = apply(Y, c(1, 2), mean, na.rm = TRUE)
colnames(mean_rt)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(mean_rt) = colnames(mean_rt)
res = plot(mean_rt,digit = 2 ,text.cell=list(pos=3, cex=0.75), col = heat.colors(7),breaks = c(200, 1000), key=NULL, axis.col = 3, cex.axis = .7)

```

### Averaging upper and lower triangles
Now we have an averaged matrix from all blocks, we average the upper and lower triangles to make the matrix symmetrical. This entails that all movements are equivalent to its reversed movement. ("101" - "1100" is equivalent to its reverse "1100" - "101")
The diagonal is forced to be 0 to make a dissimilarity matrix. Also, we are not interested in the rt between the same finger movement.
```{r,fig.width= 8, fig.height= 8}
#averaging upper and lower triangles. Then turn into array
averagedvector =  vector("list", participant_n)
for(n in 1:participant_n){
  for(i in 1:91){
    averagedvector[[n]][i]= mean(c(mean_rt[lower.tri(mean_rt)][i],t(mean_rt)[lower.tri(mean_rt)][i]),na.rm=T)
  }
}
averagedmat = matrix(, nrow =14, ncol=14)
colnames(averagedmat)= c(100,11,1010,1000,1101,101,1110,111,1011,1100,1,1001,110,10)
rownames(averagedmat) = colnames(averagedmat)
averagedreducedlist = vector("list", participant_n)
for(n in 1:participant_n){
  averagedmat[lower.tri(averagedmat)] = averagedvector[[n]]
  averagedmat = t(averagedmat)
  averagedmat[lower.tri(averagedmat)] = averagedvector[[n]]
  diag(averagedmat) = 0
  averagedreducedlist[[n]]=averagedmat
}
res = plot(averagedreducedlist[[1]],digit = 2 ,text.cell=list(pos=3, cex=0.75), col = heat.colors(7),breaks = c(200, 1000), key=NULL, axis.col = 3, cex.axis = .7)
```


